A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
intro stuff

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cycles.
The procedure is complete for ground problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
modern SMT solver. An evaluation based on problems generated from Isabelle
theories demonstrates the potential of the procedure for both proving and model
finding.
\end{abstract}

@@

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely crucially on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes are freely generated by their constructors, but unlike for datatypes
infinite constructor terms are also legitimate values
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all finite and infinite ground constructor terms, and
only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that faithfully
models the natural numbers extended with an
infinity $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides share the same infinite expansion
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
codatatype value.

@@

In this paper, we introduce a unified decision procedure for ground problems
involving both datatypes and codatatypes. The procedure is described abstractly
as a calculus (Section~\ref{sec:the-calculus}) and is design to be composable
using the Nelson--Oppen method \cite{nelson-oppen-1979}. To our knowledge, this
is the first such decision procedure for codatatypes. For the datatype case, it
roughly follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. The main differences are as follows:
%
\begin{itemize}
\item
\emph{No well-foundedness check is necessary for codatatypes.}
For example, the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed; the corresponding datatype would be rejected as
non-well-founded.

\cheat
\item
\emph{A bisimilarity rule takes the place of the cyclicity rule of datatypes.}
A cyclic constraint such as
$\const{x} \teq \const{S}(\const{x})$ %, where $\const{C}$ is a constructor,
is unsatisfiable for a datatype but satisfiable for a codatatype.
The bisimilarity principle states that two values having the same (possibly
infinite) expansion must be equal. From $\const{x} \teq \const{S}(\const{y})$ and
$\const{y} \teq \const{S}(\const{x})$, it deduces $\const{x} \teq \const{y}$.

\cheat
\item
\emph{It must be possible to express cyclic values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$ notation is used to represent cyclic ($\omega$-regular)
values. For example,
$\const{SCons}(3,\, \mu s.\; \const{SCons}(1,\, \const{SCons}(4,\, s)))$
represents the lasso-shaped sequence $3, 1, 4, 1, 4, \ldots$\kern0.1em. %%% TYPESETTING: kern
\end{itemize}

Both cyclicity and bisimilarity checks cannot be replaced by finite
axiomatizations, so they really belong in a decision procedure. They are
necessary for completeness of ground problems and for soundness of found models.
%    * in particular, acyclicity and bisimilarity are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
 And the other
(co)datatype properties, such as distinctness of constructors and
exhaustiveness, can be done more efficiently by a decision procedure than by an
axiom.
%
The proofs or soundness and completeness are included in the technical report
associated with this paper.

The procedure is implemented in the SMT solver CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the 1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using Isabelle's Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\paragraph{Conventions.}
This paper assumes a monomorphic (or many-sorted) first-order logic throughout.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set $\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values $\{\bot, \top\}$. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$), and it is treated as a logical symbol,
just like the connectives and quantifiers of first-order logic.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Upper case names (e.g., $\const{C}$,
$\const{Nil}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$
(which collapses to $\const{f} : \tau$ if $n = 0$).
For a term $t$, the notation $t : \tau$ indicates that $t$ has type $\tau$.
Functions are applied the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 : \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$,
and $\bigwedge_i \varphi_i$ abbeviates a conjunction $\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.

>>>>

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cycles.
The procedure is complete for universal problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
state-of-the-art SMT solver. An evaluation based on problems generated from
Isabelle theories demonstrates the potential of the procedure for both proving
and model finding.
\end{abstract}

@@

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have an important
role to play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely crucially on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values of codatatypes
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all well-typed finite and infinite ground constructor
terms, and only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that faithfully
models the natural numbers extended with an
infinity $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
codatatype value.

@@

In this paper, we introduce a unified decision procedure for ground problems
involving both datatypes and codatatypes. The procedure is described abstractly
as a calculus (Section~\ref{sec:the-calculus}) and is designed to be composable
via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it
roughly follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, ours is the first decision procedure for codatatypes.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. There are, however, a number of important
differences:
%
\begin{itemize}
\item
\emph{No well-foundedness check is necessary for codatatypes.}
For example, the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed; the corresponding datatype would be rejected as
non-well-founded.

\cheat
\item
\emph{A bisimilarity rule takes the place of the cyclicity rule of datatypes.}
Cyclic constraints such as
$\const{x} \teq \const{S}(\const{x})$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes but satisfiable for codatatypes.
The bisimilarity principle states that two values having the same (possibly
infinite) expansion must be equal; from $\const{x} \teq \const{S}(\const{y})$ and
$\const{y} \teq \const{S}(\const{x})$, it deduces $\const{x} \teq \const{y}$.

\cheat
\item
\emph{It must be possible to express cyclic values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic ($\omega$-regular)
values internally. For example,
the term $\const{SCons}(3,\, \mu s.\; \const{SCons}(1,\, \const{SCons}(4,\, s)))$
denotes the lasso-shaped sequence $3, 1, 4, 1, 4, \ldots$\kern0.1em. %%% TYPESETTING: kern
\end{itemize}

The cyclicity and bisimilarity rules are
needed to ensure completeness on ground problems and the absence of spurious
models in the context of finite model finding.
They cannot be replaced by finite
axiomatizations, so they naturally belong in a decision procedure. 
%    * in particular, acyclicity and bisimilarity are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
For the other (co)datatype properties---the injectivity, distinctness, and
exhaustiveness of constructors and the selector laws---it is
widely recognized that decision procedures are more efficient than
axiomatizations.
%axiom.

The proofs or soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
The procedure is implemented in the SMT solver CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\paragraph{Conventions.}
The setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values $\{\bot, \top\}$. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$) and is considered a logical symbol.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Names starting with an uppercase
letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$,
which collapses to $\const{f} : \tau$ if $n = 0$.
For a term $t$, the notation $t : \tau$ indicates that $t$ has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 : \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.
% and $\bigwedge_i \varphi_i$ abbeviates a conjunction $\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.
